Informing policy via dynamic models: Cholera in Haiti

### 2023-01-04 ms.Rnw
Implemented changes from the break

%%%% parameters %%%%%%%%%%%
\newcommand\Wsat{W_{\mathrm{sat}}}
\newcommand\muIR{\mu_{IR}}
\newcommand\muEI{\mu_{EI}}
\newcommand\transmission{\beta}
\newcommand\seasAmplitude{a}
\newcommand\rainfallExponent{r}
\newcommand\muRS{\mu_{RS}}
\newcommand\vaccineEfficacy{\theta}
\newcommand\muBirth{\mu_S}
\newcommand\muDeath{\delta}
\newcommand\choleraDeath{\delta_{C}}
\newcommand\symptomFrac{f}
\newcommand\asymptomRelativeInfect{\epsilon}
\newcommand\asymptomRelativeShed{\epsilon_{W}}
\newcommand\Wbeta[1]{\beta_{W#1}}
\newcommand\Whur[1]{\beta_{W#1}^{hm}}
\newcommand\hHur[1]{h_{#1}^{hm}}
\newcommand\tHur{t_{hm}}
\newcommand\Wremoval{\delta_W}
\newcommand\Wshed{\mu_W}
\newcommand\mixExponent{\nu}
\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
\newcommand\reportRate{\rho}
\newcommand\obsOverdispersion{\psi}
\newcommand\phaseParm{\phi}
\newcommand\transmissionTrend{\zeta}
\newcommand\Binit{\xi}
\newcommand\vaccClass{Z}
\newcommand\vaccCounter{z}
\newcommand\modelCounter{m}
\newcommand\missing{---}
\newcommand\fixed{\color{blue}}
\newcommand\demography{D}
\newcommand\code[1]{\texttt{#1}}
\newcommand\paramVec{\theta}

@@

Public health decisions must be made about when and how to implement interventions to control an infectious disease epidemic.
These decisions should be informed by data on the epidemic as well as current understanding about the transmission dynamics.
Such decisions can be posed as statistical questions about scientifically motivated dynamic models.
Thus, we encounter the methodological task of building credible, data-informed decisions based on stochastic, partially observed, nonlinear dynamic models.
This necessitates addressing the tradeoff between biological fidelity and model simplicity, and the reality of misspecification for models at all levels of complexity.
As a case study, we consider a cholera epidemic in Haiti.
The 2010 introduction of cholera to Haiti led to an extensive outbreak and sustained transmission.
Cholera was considered eliminated in 2019, though it reemerged in 2022.
We study three dynamic models developed by expert teams to advise on vaccination policies.
We assess previous methods used for fitting and evaluating these models, and we develop data analysis strategies leading to improved statistical fit.
Specifically, we present approaches to diagnosis of model misspecification, development of alternative models, and computational improvements in optimization, in the context of likelihood-based inference on nonlinear dynamic systems.
Our workflow is reproducible and extendable, facilitating future investigations of this disease system.
\end{abstract}

@@

This case study investigates the use of dynamic models and spatiotemporal data to inform a policy decision in the context of the cholera outbreak in Haiti, which started in 2010.
We build on a multi-group modeling exercise by \citet{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.
Model~1 is stochastic and describes cholera at the national level;
Model~2 is deterministic with spatial structure, and includes transmission via contaminated water;
Model~3 is stochastic with spatial structure, and accounts for measured rainfall.
Model~4 has an agent-based construction, featuring considerable mechanistic detail but limited ability to calibrate these details to data.
We focus on Models~1--3, as the strengths and weaknesses of the agent-based modeling approach  \citep{tracy18} are outside the scope of this article, and we focus on Models~1--3.

@@

\section{Mechanistic models for cholera in Haiti}\label{sec:models}
Models that focus on learning relationships between variables in a dataset are called {\it associative}, whereas models that incorporate a known scientific property of the system are called {\it causal} or {\it mechanistic}.
The danger in using forecasting techniques which rely on associative models to predict the consequence of interventions is called the Lucas critique in an econometric context.
\citet{lucas76} pointed out that it is naive to predict the effects of an intervention on a given system based entirely on historical associations.
To successfully predict the effect of an intervention, a model should therefore both provide a quantitative explanation of existing data and should have a causal interpretation: a manipulation of the system should correspond quantitatively with the corresponding change to the model.
This motivates the development of mechanistic statistical models, which provides a statistical fit to the available data while also supporting a causal interpretation.

The four mechanistic models of \citet{lee20} were deliberately developed with limited coordination.
This allows us to treat the models as fairly independently developed expert approaches to understanding cholera transmission.
However, it led to differences in notation, and in subsets of the data chosen for analysis, that hinder direct comparison.
Here, we have put all three models into a common notational framework.
Translations back to the original notation of \citet{lee20} are given in Table~S-1.

Each model describes the cholera dynamics as a partially observed Markov process (POMP) with a latent state vector $\bm{X}^{(\modelCounter)}(t)$ for each continuous time point $t$ and model $\modelCounter \in \{1, 2, 3\}$.
$N$ observations on the system are collected at time points $t_1,\dots,t_N$, written as $t_{1:N}$.
The observation at time $t_n$ is modeled by the random vector $\bm{Y}^{(\modelCounter)}_n$.
While the latent process exists between observation times, the value of the latent state at observations times is of particular interest.
We therefore write $\bm{X}^{(\modelCounter)}_n = \bm{X}^{(\modelCounter)}(t_n)$ to denote the value of the latent process at the $n$th observation time, and $\bm{X}^{(\modelCounter)}_{1:N}$ is the collection of latent state values for all observed time points.
The observable random variables $\bm{Y}^{(\modelCounter)}_{1:N}$ are assumed to be conditionally independent given $\bm{X}^{(\modelCounter)}_{0:N}$.
Together, with the density for the initial value of the latent state $\bm{X}^{(\modelCounter)}_{0} = \bm{X}^{(\modelCounter)}(t_0)$, each model assumes a joint density $f_{\bm{X}^{(\modelCounter)}_{0:N}, \bm{Y}^{(\modelCounter)}_{1:N}}\big(\bm{x}^{(\modelCounter)}_{0:N}, \bm{y}^{(\modelCounter)}_{1:N};\paramVec\big)$, where $\paramVec$ is a parameter vector that indexes the model. \arc{(Perhaps have this next sentence later when $U$ has been defined?)}
The observed data $\bm{y}_{1:N}^* = y_{1:N,1:U}^*$, along with the unobserved true value of the latent state, are modeled as a realization of this joint distribution.

For each model $\modelCounter \in \{1, 2, 3\}$, the latent state vector $\bm{X}^{(\modelCounter)}(t)$ consists of individuals labeled as susceptible (S), infected (I), asymptomatically infected (A), vaccinated (V), and recovered (R), with various sub-divisions sometimes considered.
Models~2 and~3 have metapopulation structure, meaning that each individual is a member of a spatial unit, denoted by a subscript $u\in \seq{1}{U}$.
Here, the spatial units are the $U=10$ Haitian administrative d\'{e}partements (henceforth anglicized as departments).

In the following subsections, descriptions of Models~1--3 are provided.
While the model description is scientifically critical, as well as necessary for transparency and reproducibility, the model details are not essential to our methodological discussions of how to diagnose and address model misspecification with the purpose of informing policy.
A first-time reader may choose to skim through the rest of this section, and return later.
Additional details about the numeric implementation of these models are provided in Secs.~S1 and S2 of the supplement.

@@

Latent states are initialized using an approximation of the instantaneous number of infected, asymptomatic, and recovered individuals at time $t_0$ by using the first week of cholera incidence data.
Specifically, we set $I_{u0}(0) = \frac{y^*_{1u}}{\reportRate(\muDeath + \choleraDeath + \muIR)}$, $A_{u0}(0) = \frac{1 - \symptomFrac}{\symptomFrac}I_{u0}(0)$, $R_{u0k} = y^*_{1u} - I_{u0}(0) - A_{u0}(0)$, and initialize $B_{u}(0)$ \arc{(define $B_{u}$?)} by enforcing the rainfall dynamics supposed by the one step transition model;
all other compartments that represent population counts are set to zero at time $t_0$.
For each unit $u$ with zero case counts at time $t_1$, this initialization scheme results in having zero individuals in the Infected and Asymptomatic compartments, as well as having no bacteria in the aquatic reservoir.
In reality, it's plausible that some bacteria or infected individuals were present in unit $u$ but simply went unreported.
As such, for departments with zero case counts in week 1, we consider estimating the number of infected individuals rather then treating this value as a constant (see supplement Sec.~S5 for more details).
Finally, reported cholera cases are assumed to come from a negative binomial measurement model with mean equal to a fraction ($\reportRate$) of individuals in each unit who develop symptoms and seek healthcare
(see Sec.~S3.3 of the supplement).

@@

We perform a statistical test to determine whether or not the data indicate the presence of a linear trend in transmissibility.
To do this, we perform a profile-likelihood search on the parameter $\transmissionTrend$ and obtain a confidence interval \arc{(confidence level?)} via a Monte Carlo Adjusted Profile (MCAP) \citep{ionides17}.
\cite{lee20} implemented Model~1 by fitting two distinct phases: an epidemic phase from October 2010 through March 2015, and an endemic phase from March 2015 onward.
We similarly allow the re-estimation of process and measurement overdispersion parameters ($\sigmaProc^2$ and $\obsOverdispersion$), and require that the latent Markov process $X(t)$ carry over from one phase into the next.
The resulting confidence interval for $\transmissionTrend$ is $(\Sexpr{myround(mcap_results$ci[1], digits = 3)}, \Sexpr{myround(mcap_results$ci[2], digits = 3)})$, with the full results displayed in Fig.~\ref{fig:betat}.
These results are suggestive that the inclusion of a trend in transmission rate improves the quantitative ability of Model~1 to describe the observed data.
The reported results for Model~1 in the remainder of this article were obtained with the inclusion of the parameter $\transmissionTrend$.

\begin{figure}[ht]
\centering
<<Beta_trend_Figure, fig.height=2.4, fig.width=3.6, fig.align='center'>>=
ggplot() +
  geom_point(data = h1_prof_res, aes(x = betat, y = logLik)) +
  geom_line(data = mcap_results$fit, aes(x = parameter, y = smoothed), col = 'blue') +
  geom_vline(xintercept = mcap_results$ci[1], linetype = 'dashed') +
  geom_vline(xintercept = mcap_results$ci[2], linetype = 'dashed') +
  geom_vline(xintercept = mcap_results$mle, col = 'blue') +
  labs(x = "Linear Trend in Transmission", y = 'Log Likelihood') +
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 10))
@
\caption{\label{fig:betat}
Monte Carlo adjusted profile of $\transmissionTrend$.
The blue curve is the profile, the blue line indicates the MLE, and the dashed lines indicate the confidence interval. \arc{level?}
}
\end{figure}

@@

% Model~3 parameters that are calibrated to data are primarily shared between each department, meaning that the estimated parameter value for department $u$ is the same value as the parameter value in department $v$, with $u, v \in \seq{1}{U}$.
% The exception to this being the parameters $\beta_{W_u}$, and $\beta_u$, which take department-specific values for each department $u$.
To avoid the approximation error present in high-dimensional models, \citet{lee20} simplified the problem of estimating the parameters of Model~3 by creating an approximate version of the model where the units are independent given the observed data.
% and fitting using the IF2 algorithm to estimate parameters separately for independent department-level models;
% the shared parameters were calibrated using the cholera incidence data from Artibonite, and the department-specific parameters ($\beta_{W_u}$ and $\beta_u$) were fit using the data from their respective department.
Reducing a spatially coupled model to individual units in this fashion requires special treatment of any interactive mechanisms between spatial units, such as found in Eq.~\myeqref{eq:model3:foi}.
% In particular, when considering a model for department $u$, the values $I_\nu(t)$ and $A_\nu(t)$, are unknown for $u \neq v$.
% To address this issue, \cite{lee20} conditioned each department level model on reported cholera cases in other departments and approximated the values of $I_\nu(t)$ and $A_\nu(t)$ using the respective reported cases.
Because the simplified, spatially-decoupled version of Model~3 implemented by \cite{lee20} relies on the observed cholera cases it is impossible to obtain forecasts for future time points from the fitted model.
Therefore, in order to obtain model forecasts, \cite{lee20} used the parameters estimates from the spatially-decoupled approximation of Model~3 to obtain forecasts using the fully coupled version of the model.
This approach of model calibration and forecasting avoids the issue of particle depletion, but may also be problematic.
One concern is that cholera dynamics in department $u$ are highly related to the dynamics in the remaining departments;
calibrating model parameters while conditioning on the observed cases in other departments may therefore lead to an over-dependence on observed cholera cases.
Another concern is that the two versions of the model are not the same, resulting in sub-optimal parameter estimates for the spatially coupled model,
as set of parameters that maximize the likelihood of the decoupled model almost certainly do not maximize the likelihood of the fully coupled model.
These two concerns may explain the unrealistic forecasts and low likelihood of the model (see Table~\ref{tab:likes}).

@@

The use of benchmarks may also be beneficial when developing models at differing spatial scales, where a direct comparison between model\ar{s} likelihoods is meaningless.
In such a case, a benchmark model can be fit to each spatial resolution being considered, and each model compared to their respective benchmark.
Large advantages (shortcomings) in model likelihood relative to the benchmark for a given spatial scale that are not present in other spatial scales may provide weak evidence for (against) the statistical fit of models across a range of spatial resolutions.

@@

While a useful practice, comparing model log-likelihoods to a suitable benchmark alone may not be sufficient to diagnosing the goodness-of-fit of a given model.
Additional techniques include the inspection of conditional log-likelihoods of each observation given the previous observations in order to understand how well the model describes each data point.
Other tools include plotting the effective sample size of each observation \citep{liu01}; plotting the values of the hidden states from simulations (supplement Sec.~5.1);
and comparing summary statistics of the observed data to simulations from the model \citep{wood10,king15}.

@@

Forecasts are an attempt to provide an accurate estimate of the future state of a system based on currently available data, together with an assessment of uncertainty.
Mechanistic forecasting models, compatible with current scientific understanding, may also provide estimates of the future effects of potential interventions.
Further, they may enable real-time testing of new scientific hypotheses \citep{lewis22}.

Recent information about a dynamic system should be more relevant for a forecast than older information.
This assertion may seem self-evident, but it is not the case for deterministic models, for which the initial conditions together with the parameters are sufficient for forecasting, and so recent data do not have special importance.
Epidemiological forecasts based on deterministic models are not uncommon in practice, despite their limitations \citep{king15}.
That may explain why \citet{lee20} chose to obtain forecasts by simulating the calibrated models forward from initial conditions.
Here, we compare with a forecast projected from the filtering distribution: to obtain a forecast using POMP model $m$ at a collection of times $t_{N+1:N+s}$, where $N$ is the index for the last available data, we simulate forward from a draw from
$f^{(m)}_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\theta)$
where $\hat\theta$ is a vector of calibrated parameters.
The decision not to do this partially explains the unsuccessful forecasts of \cite{lee20}: their Table~S7 shows that the subset of their simulations which were consistent with observing zero cases in 2019 predicted the elimination of cholera.
In this case study, projecting the future state of the cholera epidemic starting from latent states that are draws from the filtering distribution allows the model-based forecasts to benefit from the fact that very few cholera cases were observed in 2018 and January 2019.

Uncertainty in just a single parameter can lead to drastically different forecasts \citep{saltelli20}.
Therefore, parameter uncertainty should also be considered when obtaining model forecasts to influence policy.
If a Bayesian technique is used for parameter estimation, a natural way to account for parameter uncertainty is to obtain simulations from the model where each simulation is obtained using parameters drawn from the estimated posterior distribution.
For frequentist inference, one possible approach is obtaining model forecasts from various values of $\theta$, where the values of $\theta$ are sampled proportionally according to their corresponding likelihoods \citep{king15}; see Sec.~S7 of the supplement material for more details.
Both of these approaches share the similarity that parameters are chosen for the forecast approximately in proportion to their corresponding value of the likelihood function, $f^{(m)}_{\bm{Y}_{1:N}}(\bm{y}_{1:N}^*; \theta)$.
In this analysis, our forecasts do not account for parameter uncertainty as our goal is to comment on the estimation and diagnosis of mechanistic models, rather than suggesting our results to be used to influence vaccination policy in Haiti.
Furthermore, we use the projections from a single point estimate to highlight a deficiency of deterministic models, which is that the only variability in model projections is a result of parameter uncertainty, which leads to over-confidence in forecasts \cite{king15}.

The primary forecasting goal of \citet{lee20} was to investigate the potential consequences of vaccination interventions on a system to inform policy.
Outcomes of their study include estimates for the probability of cholera elimination and cumulative number of cholera infections under several possible vaccination scenarios.
Mimicking their efforts, we define cholera elimination as having less than one infection of cholera over at least 52 consecutive weeks, and provide forecasts under the following vaccination scenarios:

@@

Simulations from probabilistic models (Models~1 and~3) represent possible trajectories of the dynamic system under the scientific assumptions of the models.
Estimates of the probability of cholera elimination can therefore be obtained as the proportion of simulations from these models that result in cholera elimination.
The results of these projections are summarized in Figs.~\ref{fig:Mod1Scenarios}--\ref{fig:elimProbs}, and suggest that cholera elimination was likely even without increased vaccination efforts---consistent with observed reality \citep{trevisin22}.

Probability of elimination estimates of this form are not meaningful for deterministic models, as the trajectory of these models only represent the mean behavior of the system rather than individual potential outcomes.
We therefore do not provide probability of elimination estimates under Model~2.
Still, trajectories obtained by Model~2 are consistent with the simulation results of Models~1 and~3, and suggest that cholera was in the process of being eliminated from Haiti.

In additional to probability of elimination estimates, we provide estimates for the cumulative number of infections under each vaccination scenario from February 2019 -- February 2024.
Notably, the median number of cumulative cholera infections under the no-vaccination scenario using Models 1 and 3 were $\Sexpr{prettyNum(round(as.numeric(h1_S0_results$mod1_V0_probs$cumInf['q50'])), big.mark = ",")}$ and $\Sexpr{prettyNum(round(as.numeric(mod3_V0_res$mod3_V0_probs$cumInf['q50'])), big.mark = ",")}$, respectively.
While there is remaining time during this projection period in which new cholera infections can be detected, up to this point our estimates are far more consistent with the observed number of reported cholera cases than the corresponding estimates of \cite{lee20}, which were approximately $400,000$ and $1,000,000$. \arc{(I don't fully understand how both of our Model 1 and 3 cumulative case estimates can be more consistent with observed cases if our Model 1 has a lower estimate than theirs and our Model 3 has a higher estimate than theirs.)}

>>>>

%%%% parameters %%%%%%%%%%%
\newcommand\Wsat{W_{\mathrm{sat}}}
\newcommand\muIR{\mu_{IR}}
\newcommand\muEI{\mu_{EI}}
\newcommand\transmission{\beta}
\newcommand\seasAmplitude{a}
\newcommand\rainfallExponent{r}
\newcommand\muRS{\mu_{RS}}
\newcommand\vaccineEfficacy{\theta}
\newcommand\muBirth{\mu_S}
\newcommand\muDeath{\delta}
\newcommand\choleraDeath{\delta_{C}}
\newcommand\symptomFrac{f}
\newcommand\asymptomRelativeInfect{\epsilon}
\newcommand\asymptomRelativeShed{\epsilon_{W}}
\newcommand\Wbeta[1]{\beta_{W#1}}
\newcommand\Whur[1]{\beta_{W#1}^{hm}}
\newcommand\hHur[1]{h_{#1}^{hm}}
\newcommand\tHur{t_{hm}}
\newcommand\Wremoval{\delta_W}
\newcommand\Wshed{\mu_W}
\newcommand\mixExponent{\nu}
\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
\newcommand\reportRate{\rho}
\newcommand\obsOverdispersion{\psi}
\newcommand\phaseParm{\phi}
\newcommand\transmissionTrend{\zeta}
\newcommand\Binit{\xi}
\newcommand\vaccClass{Z}
\newcommand\vaccCounter{z}
\newcommand\modelCounter{m}
\newcommand\missing{---}
\newcommand\fixed{\color{blue}}
\newcommand\demography{\bullet}
\newcommand\code[1]{\texttt{#1}}
\newcommand\paramVec{\theta}

@@

Public health decisions must be made about when and how to implement interventions to control an infectious disease epidemic.
These decisions should be informed by data on the epidemic as well as current understanding about the transmission dynamics.
Such decisions can be posed as statistical questions about scientifically motivated dynamic models.
Thus, we encounter the methodological task of building credible, data-informed decisions based on stochastic, partially observed, nonlinear dynamic models.
This necessitates addressing the tradeoff between biological fidelity and model simplicity, and the reality of misspecification for models at all levels of complexity.
As a case study, we consider the 2010-2019 cholera epidemic in Haiti.
We study three dynamic models developed by expert teams to advise on vaccination policies.
We assess previous methods used for fitting and evaluating these models, and we develop data analysis strategies leading to improved statistical fit.
Specifically, we present approaches to diagnosis of model misspecification, development of alternative models, and computational improvements in optimization, in the context of likelihood-based inference on nonlinear dynamic systems.
Our workflow is reproducible and extendable, facilitating future investigations of this disease system.
\end{abstract}

@@

This case study investigates the use of dynamic models and spatiotemporal data to inform a policy decision in the context of the cholera outbreak in Haiti, which started in 2010.
We build on a multi-group modeling exercise by \citet{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.
Model~1 is stochastic and describes cholera at the national level;
Model~2 is deterministic with spatial structure, and includes transmission via contaminated water;
Model~3 is stochastic with spatial structure, and accounts for measured rainfall.
Model~4 has an agent-based construction, featuring considerable mechanistic detail but limited ability to calibrate these details to data.
We focus on Models~1--3, as the strengths and weaknesses of the agent-based modeling approach  \citep{tracy18} are outside the scope of this article.

@@

\section{Mechanistic models for cholera in Haiti}\label{sec:models}
Models that focus on learning relationships between variables in a dataset are called {\it associative}, whereas models that incorporate a known scientific property of the system are called {\it causal} or {\it mechanistic}.
The danger in using forecasting techniques which rely on associative models to predict the consequence of interventions is called the Lucas critique in an econometric context.
\citet{lucas76} pointed out that it is naive to predict the effects of an intervention on a given system based entirely on historical associations.
To successfully predict the effect of an intervention, a model should therefore both provide a quantitative explanation of existing data and should have a causal interpretation: a manipulation of the system should correspond quantitatively with the corresponding change to the model.
This motivates the development of mechanistic models, which provides a statistical fit to the available data while also supporting a causal interpretation.

The four mechanistic models of \citet{lee20} were deliberately developed with limited coordination.
This allows us to treat the models as fairly independently developed expert approaches to understanding cholera transmission.
However, it led to differences in notation, and in subsets of the data chosen for analysis, that hinder direct comparison.
Here, we have put all three models into a common notational framework.
Translations back to the original notation of \citet{lee20} are given in Table~S-1 of the supplement.

Each model describes the cholera dynamics as a partially observed Markov process (POMP) with a latent state vector $\bm{X}^{(\modelCounter)}(t)$ for each continuous time point $t$ and model $\modelCounter \in \{1, 2, 3\}$.
$N$ observations on the system are collected at time points $t_1,\dots,t_N$, written as $t_{1:N}$.
The observation at time $t_n$ is modeled by the random vector $\bm{Y}^{(\modelCounter)}_n$.
While the latent process exists between observation times, the value of the latent state at observations times is of particular interest.
We therefore write $\bm{X}^{(\modelCounter)}_n = \bm{X}^{(\modelCounter)}(t_n)$ to denote the value of the latent process at the $n$th observation time, and $\bm{X}^{(\modelCounter)}_{1:N}$ is the collection of latent state values for all observed time points.
The observable random variables $\bm{Y}^{(\modelCounter)}_{1:N}$ are assumed to be conditionally independent given $\bm{X}^{(\modelCounter)}_{0:N}$.
Together, with the density for the initial value of the latent state $\bm{X}^{(\modelCounter)}_{0} = \bm{X}^{(\modelCounter)}(t_0)$, each model assumes a joint density $f_{\bm{X}^{(\modelCounter)}_{0:N}, \bm{Y}^{(\modelCounter)}_{1:N}}\big(\bm{x}^{(\modelCounter)}_{0:N}, \bm{y}^{(\modelCounter)}_{1:N};\paramVec\big)$, where $\paramVec$ is a parameter vector that indexes the model.
The observed data $\bm{y}_{1:N}^*$, along with the unobserved true value of the latent state, are modeled as a realization of this joint distribution.

For each model $\modelCounter \in \{1, 2, 3\}$, the latent state vector $\bm{X}^{(\modelCounter)}(t)$ consists of individuals labeled as susceptible (S), infected (I), asymptomatically infected (A), vaccinated (V), and recovered (R), with various sub-divisions sometimes considered.
Models~2 and~3 have metapopulation structure, meaning that each individual is a member of a spatial unit, denoted by a subscript $u\in \seq{1}{U}$, in which case we denote the observed data for each unit using $\bm{y}_{1:N}^* = y_{1:N,1:U}^*$.
Here, the spatial units are the $U=10$ Haitian administrative d\'{e}partements (henceforth anglicized as departments).

In the following subsections, descriptions of Models~1--3 are provided.
While the model description is scientifically critical, as well as necessary for transparency and reproducibility, the model details are not essential to our methodological discussions of how to diagnose and address model misspecification with the purpose of informing policy.
A first-time reader may choose to skim through the rest of this section, and return later.
Additional details about the numeric implementation of these models are provided in Sec.~S1 and S2 of the supplement.

@@

Latent states are initialized using an approximation of the instantaneous number of infected, asymptomatic, and recovered individuals at time $t_0$ by using the first week of cholera incidence data.
Specifically, we set $I_{u0}(0) = \frac{y^*_{1u}}{\reportRate(\muDeath + \choleraDeath + \muIR)}$, $A_{u0}(0) = \frac{1 - \symptomFrac}{\symptomFrac}I_{u0}(0)$, $R_{u0k} = y^*_{1u} - I_{u0}(0) - A_{u0}(0)$, and initialize $W_{u}(0)$ by enforcing the rainfall dynamics supposed by the one step transition model;
all other compartments that represent population counts are set to zero at time $t_0$.
For each unit $u$ with zero case counts at time $t_1$, this initialization scheme results in having zero individuals in the Infected and Asymptomatic compartments, as well as having no bacteria in the aquatic reservoir.
In reality, it's plausible that some bacteria or infected individuals were present in unit $u$ but simply went unreported.
As such, for departments with zero case counts in week 1, we consider estimating the number of infected individuals rather then treating this value as a constant (see supplement Sec.~S5 for more details).
Finally, reported cholera cases are assumed to come from a negative binomial measurement model with mean equal to a fraction ($\reportRate$) of individuals in each unit who develop symptoms and seek healthcare
(see Sec.~S3.3 of the supplement).

@@

We perform a statistical test to determine whether or not the data indicate the presence of a linear trend in transmissibility.
To do this, we perform a profile-likelihood search on the parameter $\transmissionTrend$ and obtain a $95\%$ confidence interval via a Monte Carlo Adjusted Profile (MCAP) \citep{ionides17}.
\cite{lee20} implemented Model~1 by fitting two distinct phases: an epidemic phase from October 2010 through March 2015, and an endemic phase from March 2015 onward.
We similarly allow the re-estimation of process and measurement overdispersion parameters ($\sigmaProc^2$ and $\obsOverdispersion$), and require that the latent Markov process $X(t)$ carry over from one phase into the next.
The resulting confidence interval for $\transmissionTrend$ is $(\Sexpr{myround(mcap_results$ci[1], digits = 3)}, \Sexpr{myround(mcap_results$ci[2], digits = 3)})$, with the full results displayed in Fig.~\ref{fig:betat}.
These results are suggestive that the inclusion of a trend in transmission rate improves the quantitative ability of Model~1 to describe the observed data.
The reported results for Model~1 in the remainder of this article were obtained with the inclusion of the parameter $\transmissionTrend$.

\begin{figure}[ht]
\centering
<<Beta_trend_Figure, fig.height=2.4, fig.width=3.6, fig.align='center'>>=
ggplot() +
  geom_point(data = h1_prof_res, aes(x = betat, y = logLik)) +
  geom_line(data = mcap_results$fit, aes(x = parameter, y = smoothed), col = 'blue') +
  geom_vline(xintercept = mcap_results$ci[1], linetype = 'dashed') +
  geom_vline(xintercept = mcap_results$ci[2], linetype = 'dashed') +
  geom_vline(xintercept = mcap_results$mle, col = 'blue') +
  labs(x = "Linear Trend in Transmission", y = 'Log Likelihood') +
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 10))
@
\caption{\label{fig:betat}
Monte Carlo adjusted profile of $\transmissionTrend$.
The blue curve is the profile, the blue line indicates the MLE, and the dashed lines indicate the $95\%$ confidence interval.
}
\end{figure}

@@

% Model~3 parameters that are calibrated to data are primarily shared between each department, meaning that the estimated parameter value for department $u$ is the same value as the parameter value in department $v$, with $u, v \in \seq{1}{U}$.
% The exception to this being the parameters $\beta_{W_u}$, and $\beta_u$, which take department-specific values for each department $u$.
To avoid the approximation error present in high-dimensional models, \citet{lee20} simplified the problem of estimating the parameters of Model~3 by creating an approximate version of the model where the units are independent given the observed data.
% and fitting using the IF2 algorithm to estimate parameters separately for independent department-level models;
% the shared parameters were calibrated using the cholera incidence data from Artibonite, and the department-specific parameters ($\beta_{W_u}$ and $\beta_u$) were fit using the data from their respective department.
Reducing a spatially coupled model to individual units in this fashion requires special treatment of any interactive mechanisms between spatial units, such as found in Eq.~\myeqref{eq:model3:foi}.
% In particular, when considering a model for department $u$, the values $I_\nu(t)$ and $A_\nu(t)$, are unknown for $u \neq v$.
% To address this issue, \cite{lee20} conditioned each department level model on reported cholera cases in other departments and approximated the values of $I_\nu(t)$ and $A_\nu(t)$ using the respective reported cases.
Because the simplified, spatially-decoupled version of Model~3 implemented by \cite{lee20} relies on the observed cholera cases, the calibrated model cannot be used to obtain forecasts.
Therefore, in order to obtain model forecasts, \cite{lee20} used the parameters estimates from the spatially-decoupled approximation of Model~3 to obtain forecasts using the fully coupled version of the model.
This approach of model calibration and forecasting avoids the issue of particle depletion, but may also be problematic.
One concern is that cholera dynamics in department $u$ are highly related to the dynamics in the remaining departments;
calibrating model parameters while conditioning on the observed cases in other departments may therefore lead to an over-dependence on observed cholera cases.
Another concern is that the two versions of the model are not the same, resulting in sub-optimal parameter estimates for the spatially coupled model,
as set of parameters that maximize the likelihood of the decoupled model almost certainly do not maximize the likelihood of the fully coupled model.
These two concerns may explain the unrealistic forecasts and low likelihood of the model (see Table~\ref{tab:likes}).

@@

The use of benchmarks may also be beneficial when developing models at differing spatial scales, where a direct comparison between model likelihoods is meaningless.
In such a case, a benchmark model can be fit to each spatial resolution being considered, and each model compared to their respective benchmark.
Large advantages (shortcomings) in model likelihood relative to the benchmark for a given spatial scale that are not present in other spatial scales may provide weak evidence for (against) the statistical fit of models across a range of spatial resolutions.

@@

While a useful practice, comparing model log-likelihoods to a suitable benchmark alone may not be sufficient to diagnosing the goodness-of-fit of a given model.
Additional techniques include the inspection of conditional log-likelihoods of each observation given the previous observations in order to understand how well the model describes each data point (Sec.~S5).
Other tools include plotting the effective sample size of each observation \citep{liu01}; plotting the values of the hidden states from simulations (supplement Sec.~5.1);
and comparing summary statistics of the observed data to simulations from the model \citep{wood10,king15}.

@@

Forecasts are an attempt to provide an accurate estimate of the future state of a system based on currently available data, together with an assessment of uncertainty.
Forecasts from mechanistic models that are compatible with current scientific understanding may also provide estimates of the future effects of potential interventions.
Further, they may enable real-time testing of new scientific hypotheses \citep{lewis22}.

Recent information about a dynamic system should be more relevant for a forecast than older information.
This assertion may seem self-evident, but it is not the case for deterministic models, for which the initial conditions together with the parameters are sufficient for forecasting, and so recent data do not have special importance.
Epidemiological forecasts based on deterministic models are not uncommon in practice, despite their limitations \citep{king15}.
That may explain why \citet{lee20} chose to obtain forecasts by simulating the calibrated models forward from initial conditions.
Here, we compare with a forecast projected from the filtering distribution: to obtain a forecast using POMP model $m$ at a collection of times $t_{N+1:N+s}$, where $N$ is the index for the last available data, we simulate forward from a draw from
$f^{(m)}_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\theta)$
where $\hat\theta$ is a vector of calibrated parameters.
The decision not to do this partially explains the unsuccessful forecasts of \cite{lee20}: their Table~S7 shows that the subset of their simulations which were consistent with observing zero cases in 2019 also accurately predicted the prolonged absence of detected cholera.

Uncertainty in just a single parameter can lead to drastically different forecasts \citep{saltelli20}.
Therefore, parameter uncertainty should also be considered when obtaining model forecasts to influence policy.
If a Bayesian technique is used for parameter estimation, a natural way to account for parameter uncertainty is to obtain simulations from the model where each simulation is obtained using parameters drawn from the estimated posterior distribution.
For frequentist inference, one possible approach is obtaining model forecasts from various values of $\theta$, where the values of $\theta$ are sampled proportionally according to their corresponding likelihoods \citep{king15}; see Sec.~S7 of the supplement material for more details.
Both of these approaches share the similarity that parameters are chosen for the forecast approximately in proportion to their corresponding value of the likelihood function, $f^{(m)}_{\bm{Y}_{1:N}}(\bm{y}_{1:N}^*; \theta)$.
In this analysis, our forecasts do not account for parameter uncertainty as our goal is to comment on the estimation and diagnosis of mechanistic models, rather than influencing vaccination policy in Haiti.
Furthermore, we use the projections from a single point estimate to highlight a deficiency of deterministic models, which is that the only variability in model projections is a result of parameter uncertainty, which leads to over-confidence in forecasts \cite{king15}.

The primary forecasting goal of \citet{lee20} was to investigate the potential consequences of vaccination interventions on a system to inform policy.
One outcome of their study include estimates for the probability of cholera elimination under several possible vaccination scenarios.
Mimicking their efforts, we define cholera elimination as having no cholera infections over at least 52 consecutive weeks, and provide forecasts under the following vaccination scenarios:

@@

Simulations from probabilistic models (Models~1 and~3) represent possible trajectories of the dynamic system under the scientific assumptions of the models.
Estimates of the probability of cholera elimination can therefore be obtained as the proportion of simulations from these models that result in cholera elimination.
The results of these projections are summarized in Figs.~\ref{fig:Mod1Scenarios}--\ref{fig:elimProbs}.

Probability of elimination estimates of this form are not meaningful for deterministic models, as the trajectory of these models only represent the mean behavior of the system rather than individual potential outcomes.
We therefore do not provide probability of elimination estimates under Model~2, but show trajectories under the various vaccination scenarios using this model (Fig.~\ref{fig:mod2Traj}).
